Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-10 csf-2017/paper.tex
pass over examples

\subsection{Encapsulation of local state}
The capability machine provides encapsulation of local state. The
programs \texttt{f1} in \figurename~\ref{fig:prog-f1} and 
\texttt{f2} \figurename~\ref{fig:prog-f2} demonstrate this.
The two programs \texttt{f1} and \texttt{f2} are very similar. They both
store some local data, call an untrusted piece of code ($\adv$), and finally
make sure that their local state is unchanged. They differ in the way
they do this. Program \texttt{f1} assumes it has access to a stack capability
and uses the stack to store its local state. It also uses the
stack-based calling convention to call the untrusted code. On the
other hand, \texttt{f2} allocates a new piece of memory for its local
state and uses an activation-record based calling convention
(described in the technical report) to run the adversarial code.

@@

For both programs, we can prove that if they are linked with an
adversary, $\adv$, that is allowed to allocate new memory but otherwise has no
capabilities, then the assertion will never fail when
executing the program,

We can only prove this, because the capability machine encapsulates
the local state of the two programs. The two examples also illustrate
the versatility of the logical relation. The two examples use
different calling convention, but the logical relation is not specific
to any calling convention, so we can use it to reason about both
programs.

% \todo[inline]{give a more detailed high-level taste of a proof here? Maybe later for one of the other examples?}
% \begin{lemma}[Correctness lemma for \texttt{f1}]
% \todo[inline]{The lemma copied directly from the TR. To be deleted later!}
%   \label{lem:correctness-f1}
%   Let
%   \begin{align*}
%     c_{\var{adv}} & \defeq ((\entry,\glob),\start_{\adv},\addrend_{\adv},\start_{\adv}+\olf) \\
%     c_{f1} & \defeq ((\rwx,\glob),\mathtt{f1}-\olf,\mathtt{1f},\mathtt{f1}) \\
%     c_\malloc & \defeq ((\entry,\glob),\start_\malloc,\addrend_\malloc,\start_\malloc+\olf) \\
%     c_{\var{stk}} & \defeq ((\rwlx,\local),\start_\stk,\addrend_\stk,\start_\stk-1) \\
%     c_\link & \defeq ((\readonly,\glob),\link,\link+1,\link)\\
%     \reg & \in \Regs \\
%     m & \defeq \hs_{f1} \uplus 
%         \hs_\flag \uplus                
%         \ms_{\var{link}} \uplus 
%         \hs_\adv \uplus 
%         \ms_{\malloc} \uplus 
%         \ms_{\var{stk}} \uplus
%         \ms_{\var{frame}} 
%   \end{align*}
%   and
%   \begin{itemize}
%   \item $c_\malloc$ satisfies the specification for malloc and $\iota_{\malloc,0}$ is the region from the specification.
%   \end{itemize}
%   where 
%   \begin{align*}
%     &\dom(\hs_{f1}) = [\mathtt{f1}-\olf,\mathtt{1f}] \\
%     &\dom(\hs_\flag) = [\flag,\flag] \\
%     &\dom(\ms_\link) = [\link,\link+1]\\
%     &\dom(\ms_\stk) = [\start_\stk, \addrend_\stk]\\
%     &\dom(\hs_{\adv}) = [\start_\adv,\addrend_\adv] \\
%     &\heapSat[\hs_{\malloc}]{n}{[0 \mapsto \iota_{\malloc,0}]} \qquad \text{ for all $n \in \nats$}
%   \end{align*}
%   and
%   \begin{itemize}
%   \item $\ms_{f1}(\mathtt{f1}-\olf) = ((\readonly,\glob),\link,\link+1,\link)$, $\ms_{f1}(\mathtt{f1}-\olf+1) = ((\readwrite,\glob),\flag,\flag,\flag)$, the rest of $\hs_{f1}$ contains the code of $f1$.
%   \item $\ms_\flag = [\flag \mapsto 0]$
%   \item $\ms_{\var{link}} = [\var{link} \mapsto c_\malloc, \var{link} + 1 \mapsto c_\adv]$
%   \item $\hs_\adv(\start_\adv) = c_\link$ and $\forall \addr \in [\start_\adv+1,\addrend]\ldotp \ms_\adv(a) \in \ints$
%   \end{itemize}
%   if 
%   \[
%     (\reg\update{\pcreg}{c_{f1}}\update{r_\stk}{c_\stk},m) \step[n] (\halted,m'),
%   \]
%   then
%   \[
%     m'(\flag) = 0
%   \]  
% \end{lemma}

In order to formulate a lemmas about \texttt{f1} and \texttt{f2}, we need a way to observe whether the assertion fails. To this end, we assum access to a flag which will be set by the assertion if it fails. A slightly simplified version of the correctness lemma for \texttt{f1} is:
\begin{lemma}
  \label{lem:correctness-f1}
  Let
  \begin{align*}
    c_{\var{adv}} & \defeq ((\entry,\glob),\dots) \\
    c_{f1} & \defeq ((\rwx,\glob),\dots) \\
    c_\malloc & \defeq ((\entry,\glob),\dots) \\
    c_{\var{stk}} & \defeq ((\rwlx,\local),\dots) \\
    c_\link & \defeq ((\readonly,\glob),\dots)\\
    \reg & \in \Regs \\
    m & \defeq
        \begin{aligned}[t]
          &\hs_{f1} \uplus 
        \hs_\flag \uplus                
        \ms_{\var{link}} \uplus 
        \hs_\adv \uplus \\
        &\ms_{\malloc} \uplus 
        \ms_{\var{stk}} \uplus
        \ms_{\var{frame}}
      \end{aligned}
  \end{align*}
  where each of the capability has an appropriate range of authority and pointer and
  \begin{itemize}
  \item $\ms_{f1}$ contains $c_\link$, $c_\flag$ and the code of \texttt{f1}
  \item $\ms_\flag(\flag) = 0$
  \item $\ms_{\var{link}}$ contains $c_\adv$ and $c_\malloc$
  \item $\hs_\adv$ contains $c_\link$ and otherwise only instructions.
  \end{itemize}
  if 
  \[
    (\reg\update{\pcreg}{c_{f1}}\update{r_\stk}{c_\stk},m) \step^* (\halted,m'),
  \]
  then
  \[
    m'(\flag) = 0
  \]  
\end{lemma}

In order for prove Lemma~\ref{lem:correctness-f1}, it suffices to show
that the start configuration is capability safe (in the $\observations$ relation) for a
world with a permanent region that requires the assertion flag to be
0. This is done using an anti-reduction lemma which states that for a
configuration to be capability safe, it is sufficient to
show that it is in it at a later point of the execution. To reason
about \texttt{scall}, it suffices to show that the configuration just
after we jumped is capability safe and that the configuration just
after \texttt{scall} is done cleaning up is capability safe.
We use the Fundamental Theorem to reason about the unknown code, but
notice that the Fundamental Theorem says nothing about enter
capabilities and that the adversary capability is an enter
capability. Luckily the enter capability becomes a read/execute
capability when we jump and the Fundamental Theorem can be used to
show that this capability is capability safe as a program counter.

>>>>

\subsection{Encapsulation of local state}
The capability machine provides encapsulation of local state. The
programs \texttt{f1} in \figurename~\ref{fig:prog-f1} and 
\texttt{f2} \figurename~\ref{fig:prog-f2} demonstrate this.
The two programs \texttt{f1} and \texttt{f2} are very similar. They both
store some local data, call an untrusted piece of code ($\adv$), and
then tests whether local state is unchanged. They differ in the way
they do this. Program \texttt{f1} assumes it has access to a stack capability
and uses the stack to store its local state. It also uses the
stack-based calling convention to call the untrusted code. On the
other hand, \texttt{f2} allocates a new piece of memory for its local
state and uses an activation-record based calling convention
(described in the technical report) to run the adversarial code.

@@

For both programs, we can prove that if they are linked with an
adversary, $\adv$, that is allowed to allocate new memory but otherwise has no
capabilities, then the assertion will never fail when
executing the program (see Lemmas~\ref{lem:correctness-f1}
and~\ref{lem:correctness-f2} below).
\lars{shouldn't we say already here that the adversary is a global 
  capability and that is why, intuitively, the local state is
  preserved ?}
The two examples also illustrate
the versatility of the logical relation. The two examples use
different calling convention, but the logical relation is not specific
to any calling convention, so we can use it to reason about both
programs.

% \begin{lemma}[Correctness lemma for \texttt{f1}]
% \todo[inline]{The lemma copied directly from the TR. To be deleted later!}
%   \label{lem:correctness-f1}
%   Let
%   \begin{align*}
%     c_{\var{adv}} & \defeq ((\entry,\glob),\start_{\adv},\addrend_{\adv},\start_{\adv}+\olf) \\
%     c_{f1} & \defeq ((\rwx,\glob),\mathtt{f1}-\olf,\mathtt{1f},\mathtt{f1}) \\
%     c_\malloc & \defeq ((\entry,\glob),\start_\malloc,\addrend_\malloc,\start_\malloc+\olf) \\
%     c_{\var{stk}} & \defeq ((\rwlx,\local),\start_\stk,\addrend_\stk,\start_\stk-1) \\
%     c_\link & \defeq ((\readonly,\glob),\link,\link+1,\link)\\
%     \reg & \in \Regs \\
%     m & \defeq \hs_{f1} \uplus 
%         \hs_\flag \uplus                
%         \ms_{\var{link}} \uplus 
%         \hs_\adv \uplus 
%         \ms_{\malloc} \uplus 
%         \ms_{\var{stk}} \uplus
%         \ms_{\var{frame}} 
%   \end{align*}
%   and
%   \begin{itemize}
%   \item $c_\malloc$ satisfies the specification for malloc and $\iota_{\malloc,0}$ is the region from the specification.
%   \end{itemize}
%   where 
%   \begin{align*}
%     &\dom(\hs_{f1}) = [\mathtt{f1}-\olf,\mathtt{1f}] \\
%     &\dom(\hs_\flag) = [\flag,\flag] \\
%     &\dom(\ms_\link) = [\link,\link+1]\\
%     &\dom(\ms_\stk) = [\start_\stk, \addrend_\stk]\\
%     &\dom(\hs_{\adv}) = [\start_\adv,\addrend_\adv] \\
%     &\heapSat[\hs_{\malloc}]{n}{[0 \mapsto \iota_{\malloc,0}]} \qquad \text{ for all $n \in \nats$}
%   \end{align*}
%   and
%   \begin{itemize}
%   \item $\ms_{f1}(\mathtt{f1}-\olf) = ((\readonly,\glob),\link,\link+1,\link)$, $\ms_{f1}(\mathtt{f1}-\olf+1) = ((\readwrite,\glob),\flag,\flag,\flag)$, the rest of $\hs_{f1}$ contains the code of $f1$.
%   \item $\ms_\flag = [\flag \mapsto 0]$
%   \item $\ms_{\var{link}} = [\var{link} \mapsto c_\malloc, \var{link} + 1 \mapsto c_\adv]$
%   \item $\hs_\adv(\start_\adv) = c_\link$ and $\forall \addr \in [\start_\adv+1,\addrend]\ldotp \ms_\adv(a) \in \ints$
%   \end{itemize}
%   if 
%   \[
%     (\reg\update{\pcreg}{c_{f1}}\update{r_\stk}{c_\stk},m) \step[n] (\halted,m'),
%   \]
%   then
%   \[
%     m'(\flag) = 0
%   \]  
% \end{lemma}

In order to formulate lemmas about \texttt{f1} and \texttt{f2}, we
need a way to observe whether the assertion fails. To this end, we
assume access to a flag (an address in memory)
which will be set to $1$ by the assertion if it
fails. A slightly simplified version of the correctness lemma for
\texttt{f1} is:
\begin{lemma}
  \label{lem:correctness-f1}
  Let
  \begin{align*}
    c_{\var{adv}} & \defeq ((\entry,\glob),\dots) \\
    c_{f1} & \defeq ((\rwx,\glob),\dots) \\
    c_\malloc & \defeq ((\entry,\glob),\dots) \\
    c_{\var{stk}} & \defeq ((\rwlx,\local),\dots) \\
    c_\link & \defeq ((\readonly,\glob),\dots)\\
    \reg & \in \Regs \\
    m & \defeq
        \begin{aligned}[t]
          &\hs_{f1} \uplus 
        \hs_\flag \uplus                
        \ms_{\var{link}} \uplus 
        \hs_\adv \uplus \\
        &\ms_{\malloc} \uplus 
        \ms_{\var{stk}} \uplus
        \ms_{\var{frame}}
      \end{aligned}
  \end{align*}
  where each of the capability has an appropriate range of authority and pointer and
  \begin{itemize}
  \item $\ms_{f1}$ contains $c_\link$, $c_\flag$ and the code of \texttt{f1}
  \item $\ms_\flag(\flag) = 0$
  \item $\ms_{\var{link}}$ contains $c_\adv$ and $c_\malloc$
  \item $\hs_\adv$ contains $c_\link$ and otherwise only instructions.
  \end{itemize}
  if 
  \[
    (\reg\update{\pcreg}{c_{f1}}\update{r_\stk}{c_\stk},m) \step^* (\halted,m'),
  \]
  then
  \[
    m'(\flag) = 0
  \]  
\end{lemma}

To prove Lemma~\ref{lem:correctness-f1}, it suffices to show
that the start configuration is capability safe (in the $\observations$ relation) for a
world with a permanent region that requires the assertion flag to be
0. This is done using an anti-reduction lemma which states that for a
configuration to be capability safe, it is sufficient to
show that it is in it at a later point of the execution. To reason
about \texttt{scall}, it suffices to show that (1) the configuration just
after we jumped is capability safe and (2) that the configuration just
after \texttt{scall} is done cleaning up is capability safe.
We use the Fundamental Theorem to reason about the unknown code, but
notice that the Fundamental Theorem says nothing about enter
capabilities and that the adversary capability is an enter
capability. Luckily the enter capability becomes a read/execute
capability when we jump and the Fundamental Theorem can be used to
show that this capability is capability safe as a program counter.
